\section{Learning Bayesian Networks}
Learning Bayesian Networks (BNs) from data involves two main tasks: parameter learning and structure learning. 
In this section, we will discuss both tasks in detail.

\subsection{Parameter Learning}
For parameter learning, we assume that the structure $G$ of the Bayesian Network is given.
We are given a dataset $D = \{\boldsymbol{x}(1), \boldsymbol{x}(2), \ldots, \boldsymbol{x}(N)\}$, 
where each $\boldsymbol{x}_i$ is a configuration of all (\textbf{complete data}) or 
some (\textbf{incomplete data}) of the variables in the network.
\\The goal is to estimate the parameters $\theta$ of the model (one parameter per variable) from the data.
Again, the simplest approach consists in finding the parameters that maximize the likelihood of the data (MLE):
\[ \theta^* = \arg\max_{\theta} P(D | \theta) = \arg\max_{\theta} \mathcal{L}(D, \theta) \]

\subsubsection{Complete Data}
Given the complete data $D$, since every datapoint is \textit{i.i.d.} (independent and identically distributed), 
we can express the likelihood of the data as:
\begin{equation}
    P(D|\theta) = \prod_{i=1}^{N} P(\boldsymbol{x}(i) | \theta)
\end{equation}
Instead of working with the entire structure $G$ for each data point, we can exploit the 
factorization property of Bayesian Networks. This allows us to express $p(\boldsymbol{x}(i) | \theta)$ as:
\begin{equation}
    P(D|\theta) = \prod_{i=1}^{N} \prod_{j=1}^{m} P(\boldsymbol{x}_{j}(i) | \text{Pa}_{j}(i),\theta)
\end{equation}
Where:
\begin{itemize}
    \item $\boldsymbol{x}_{j}(i)$ is the value of variable $X_j$ in the $i$-th data point;
    \item $\text{Pa}_{j}(i)$ is the set of values of the parent variables of $X_j$ in the $i$-th data point.
\end{itemize}
Graphically, we can represent the likelihood as follows:
\begin{figure}[H]
\centering
    \begin{tikzpicture}[
      bn-node/.style={circle, draw=red, thick, minimum size=10pt, inner sep=1pt, font=\scriptsize}
    ]
    % Theta nodes
    \node[bn-theta] (TC) at (0, 2.8) {$\theta_1$};
    \node[bn-theta] (TA) at (-2, -1.4) {$\theta_{2}|1$};
    \node[bn-theta] (TB) at (2, -1.4) {$\theta_{3}|1$};
    % Nodes
    \node[bn-node] (C1) at (-4, 1.4) {$X_1(1)$};
    \node[bn-node] (A1) at (-5, 0) {$X_2(1)$};
    \node[bn-node] (B1) at (-3, 0) {$X_3(1)$};

    \node[bn-node] (C2) at (0, 1.4) {$X_1(2)$};
    \node[bn-node] (A2) at (-1, 0) {$X_2(2)$};
    \node[bn-node] (B2) at (1, 0) {$X_3(2)$};

    \node[bn-node] (CN) at (4, 1.4) {$X_1(N)$};
    \node[bn-node] (AN) at (5, 0) {$X_2(N)$};
    \node[bn-node] (BN) at (3, 0) {$X_3(N)$};

    % Edges
    \draw[bn-arrow] (C1) -- (A1);
    \draw[bn-arrow] (C1) -- (B1);
    \draw[bn-arrow] (C2) -- (A2);
    \draw[bn-arrow] (C2) -- (B2);
    \draw[bn-arrow] (CN) -- (AN);
    \draw[bn-arrow] (CN) -- (BN);
    
    % Theta edges
    \draw[bn-arrow] (TC) -- (C1);
    \draw[bn-arrow] (TC) -- (C2);
    \draw[bn-arrow] (TC) -- (CN);
    \draw[bn-arrow] (TA) -- (A1);
    \draw[bn-arrow] (TA) -- (A2);
    \draw[bn-arrow] (TA) -- (AN);
    \draw[bn-arrow] (TB) -- (B1);
    \draw[bn-arrow] (TB) -- (B2);
    \draw[bn-arrow] (TB) -- (BN);
    \end{tikzpicture}
    \caption{Complete data representation}
\end{figure}
The probability $p(\boldsymbol{x}_j(i) | pa_j(i), \theta)$ only depends on the parameters $\theta_{X_j|Pa_j}$,
not on the parameters of other variables.
Because the Conditional Probability Distributions (CPDs) are independent, we can rearrange the likelihood as:
\begin{equation}
    P(D|\theta) = \prod_{j=1}^{m} \prod_{i=1}^{N} P(x_j(i) | pa_j(i), \theta_{X_j|Pa_j})
\end{equation}
This allows us to decompose the MLE problem into $m$ independent subproblem, one for each variable $X_j$:
\begin{equation}
    \label{eq:mle-subproblem}
    \theta_{X_j|Pa_j}^{\max} = \arg\max_{\theta_{X_j|Pa_j}}\underbrace{\prod_{i=1}^{N} P(x_j(i) | pa_j(i), \theta_{X_j|Pa_j})}_{\mathcal{L}(\theta_{X_j|Pa_j, \mathcal{D}})}
\end{equation}
When working with discrete variables, we can estimate the parameters using frequency counts from the data.
A discrete CPD $P(X | U)$ can be represented as a table:
\begin{itemize}
    \item Each row corresponds to a specific configuration of $X$;
    \item Each column corresponds to a specific configuration of the parent variables $U$;
    \item Each table entry $\theta_{x|u}$ represents the probability of the configuration $X = x$ and $U = u$.
\end{itemize}
From equation \ref{eq:mle-subproblem}, we can replace $P(x_j(i) | pa_j(i))$ with the corresponding parameter 
$\theta_{x(i)|u(i)}$. The likelihood for a single CPD $X$ becomes:
\begin{equation*}
    \mathcal{L}(\theta_{X|U}, D) = \prod_{i=1}^{N} \theta_{x(i)|u(i)}
\end{equation*}
\begin{equation*}
    = \prod_{u \in Val(U)} \left[\prod_{x \in Val(X)} \theta_{x|u}^{N_{x,u}}\right]
\end{equation*}
Where $N_{x,u}$ is the count of occurrences in the data where $X = x$ and $U = u$.
Now, each column in the CPD contains a multinomial distribution over the values of $X$, 
given a specific configuration of its parents $U = u$.
This implies that each column must sum to 1:
\[\sum_{x \in Val(X)} \theta_{x|u} = 1\]
Parameters of different columns are independent, allowing us to estimate them independently.
Thus, we can decompose the MLE problem for each column $u$ as follows:
\begin{equation*}
    \theta_{x|u}^{\max} = \frac{N_{u,x}}{\sum_{x} N_{u,x}}
\end{equation*}
This means that the MLE estimate for each parameter $\theta_{x|u}$ is simply the relative frequency of $X = x$
given $U = u$ in the data.

\subsubsection{Adding priors}
The maximum likelihood estimation can lead to overfitting on the training data, especially when the dataset is small.
Also, configuration of variables that do not appear in the training data will have zero probability.
A common solution to these problems is to introduce a prior distribution over the parameters $\theta$,
based on prior knowledge or assumptions.
This way, we now want to maximize the posterior distribution of the parameters given the data:
\begin{equation*}
    \theta^* = \arg\max_{\theta} P(\theta | D)
\end{equation*}
Applying Bayes' theorem, we can rewrite this as:
\begin{equation*}
    P(\theta | D) =\frac{P(D | \theta) P(\theta)}{P(D)}
\end{equation*}
Since $P(D)$ is constant with respect to $\theta$, we can simplify the optimization problem to:
\begin{equation}
    \theta^* = \arg\max_{\theta} P(D | \theta) P(\theta)
\end{equation}
This approach is known as Maximum A Posteriori (MAP) estimation.
\\The conjugate prior for the multinomial distribution is the Dirichlet distribution, 
with parameters $\alpha_{x|u}$, for each possible value of $x$.
The resulting MAP estimate for each parameter $\theta_{x|u}$ becomes:
\begin{equation}*
    \theta_{x|u}^{\max} = \frac{N_{u,x} + \alpha_{x|u}}{\sum_{x} (N_{u,x} + \alpha_{x|u})}
\end{equation}
Where $\alpha_{x|u}$ can be interpreted as imaginary samples with configuration $X_j = x$ and $Pa_j = u$.

\subsubsection{Incomplete Data}
With incomplete data, some examples may have missing values for certain variables.
In such cases, counts $N_{x|u}$ cannot be directly computed from the data.
To address this, we can use an approximate method called the Expectation-Maximization (EM) algorithm.
\paragraph{EM algorithm}
In a nutshell, as we have said before, sufficient statistics (like counts) are not directly computable from incomplete data.
The EM algorithm iteratively estimates these statistics based on the current parameter estimates, 
and then updates the parameters using these estimated statistics.
\\The algorithm consists of two main steps:
\begin{itemize}
    \item \textbf{Expectation step (E-step)}: compute the expected sufficient statistics for the complete data,
    with respect to the joint distribution for $\boldsymbol{X}$ conditioned on the current
    value of $\theta$ (current estimates of the parameters) and the known data $\mathcal{D}$:
    \begin{equation}
        E[N_{ijk}] = \sum_{l=1}^{n} P(X_i(l) = x_{k}, Pa_i(l) = pa_{j} | \boldsymbol{X}_l, \theta)
    \end{equation}
    Where:
    \begin{itemize}
        \item $N_{ijk}$ is the count of occurrences where variable $X_i$ takes value $x_k$
        and its parents $Pa_i$ take configuration $pa_j$;
        \item $\boldsymbol{X}_l$ is the $l$-th data point in the dataset $\mathcal{D}$ (observed values);
        \item if $X_i(l)$ and $Pa_i(l)$ are observed for $X_l$ in the dataset $\mathcal{D}$, then the probability is either 0 or 1;
        \item Otherwise, we run Bayesian inference to compute the probability.
    \end{itemize}
    \item \textbf{Maximization step (M-step)}: compute the parameters maximizing likelihood of the
    complete dataset $D_c$, using the expected counts from the E-step:
    \begin{equation}
        \theta_{ijk}^{\max} = \frac{E[N_{ijk}]}{\sum_{k} E[N_{ijk}]}
    \end{equation}
    This is the same formula as in the complete data case, but using expected counts instead of actual counts.
\end{itemize}
As previously seen in the complete data case, also in this case we can use a MAP estimation approach by 
introducing a prior distribution over the parameters. This is not different from the complete data case,
for the only exception being the use of expected counts instead of actual counts.

\subsection{Structure Learning}
There are different approaches to learn the structure of a Bayesian Network from data.
Some of this approaches include:
\begin{itemize}
    \item \textbf{Constraint-based methods}: these methods use statistical tests to 
    identify conditional independencies between variables. 
    Based on these independencies, a graph structure is proposed.
    \item \textbf{Score-based methods}: these methods define a scoring for each possible structure based on how well it fits the data.
    It then searches the space of possible structures to find the one that maximizes the score.
    \item \textbf{Model-averaging methods}: these methods consider multiple possible structures 
    and average their predictions, weighted by their probabilities.
\end{itemize}