\section{Linear Discriminant Functions}
Before deep diving into specific learning algorithms, it is important to understand the difference between 
\textbf{generative models} and \textbf{discriminative models}.
\begin{itemize}
    \item \textbf{Generative models} aim to model the \textbf{distribution} governing the data. 
    They learn the joint probability distribution $P(x,y)$ and 
    can generate new data points, according to the learned distribution.
    Some examples of generative models are \textbf{Naive Bayes}, \textbf{Gaussian Mixture Models}.
    \item \textbf{Discriminative models} aim to model the \textbf{discriminative function} directly, 
    rather than the underlying data distribution. For example, for classification tasks,
    they directly model the \textbf{decision boundary} that separates different classes.
\end{itemize}

\subsection{Advantages and disadvantages of discriminative models}
\paragraph{Pros:}
\begin{itemize}
    \item When data is complex, modelling the underlying distribution can be very challenging 
    (E.g., high resolution images);
    \item If \textbf{data discrimination} is the final goal, then there is no need to model the entire data distribution;
    \item Discriminative models can focus on learning the parameters that are most relevant for learning 
    the decision boundary, potentially leading to better performance.
\end{itemize}
\paragraph{Cons:}
\begin{itemize}
    \item The learned model is less flexible in its usage;
    \item It doesn't allow to perform arbitrary inference tasks on the data. 
    In example, it is not possible to generate new data points from the learned model.
\end{itemize}

\subsection{Definition}
\label{sec:linear_discriminant_functions}
A \textbf{linear discriminant model} uses a \textbf{linear function} to make predictions. 
The \textbf{decision boundary} (discriminant function) is a linear combination of the input features:
\begin{equation} \label{eq:linear_discriminant}
    f(\boldsymbol{x}) = \boldsymbol{w}^T \boldsymbol{x} + w_0
\end{equation}
where:
\begin{itemize}
    \item $\boldsymbol{x}$ is the input feature vector;
    \item $\boldsymbol{w}$ is the weight vector, representing the importance of each feature;
    \item $w_0$ is the bias term, which allows to shift the decision boundary.
\end{itemize}
Multiplying $\boldsymbol{w}^T$ by $\boldsymbol{x}$ computes a weighted sum of the input features,
and adding $w_0$ shifts the result.
\\The linear discriminant function is the simplest form of discriminant function. 
Depending on the specific problem and data, more complex forms of discriminant functions can be used,
but linear functions should always be considered as a baseline.

\subsection{Linear Binary Classifier}
In a \textbf{binary classification} problem, the linear discriminant function can be used to classify data points into two classes.
Since the basic output of the discriminant function is a continuous value, we need to define a threshold to convert it into a binary prediction.
This is usually done by taking the sign of the discriminant function:
\begin{equation}
    f(\boldsymbol{x}) = \text{sign}(\boldsymbol{w}^T \boldsymbol{x} + w_0) = 
    \begin{cases}
    +1 & \text{if } \boldsymbol{w}^T \boldsymbol{x} + w_0 \geq 0 \\
    -1 & \text{if } \boldsymbol{w}^T \boldsymbol{x} + w_0 < 0
    \end{cases}
\end{equation}
The \textbf{decision boundary}, which is a \textbf{hyperplane} $H$ in the feature space, is defined by the equation:
\begin{equation}
    f(\boldsymbol{x}) = 0
\end{equation}
The weight vector $\boldsymbol{w}$ is \textbf{orthogonal} to the decision boundary $H$.

\begin{proof}
    We take two points on the decision boundary, $\boldsymbol{x}_1$ and $\boldsymbol{x}_2$, such that:
    \[ f(\boldsymbol{x}_1) = f(\boldsymbol{x}_2) = 0 \]
    Based on equation \eqref{eq:linear_discriminant}, we can rewrite this as:
    \[ \boldsymbol{w}^T \boldsymbol{x}_1 + w_0 = \boldsymbol{w}^T \boldsymbol{x}_2 + w_0 = 0 \]
    \[ \boldsymbol{w}^T \boldsymbol{x}_1 - \boldsymbol{w}^T \boldsymbol{x}_2 = 0 \]
    \[ \boldsymbol{w}^T (\boldsymbol{x}_1 - \boldsymbol{x}_2) = 0 \]
    This shows that the vector $\boldsymbol{w}^T$ is orthogonal to the vector $(\boldsymbol{x}_1 - \boldsymbol{x}_2)$, which lies on the decision boundary. 
\end{proof}
\definition{Functional margin}{
    The value $f(\boldsymbol{x})$ of the discriminative function for a certain point $\boldsymbol{x}$
    is called \textbf{functional margin}. 
    It can be seen as the \textbf{confidence} of the model in the prediction for that point. 
    A larger absolute value of the functional margin indicates a higher confidence in the prediction.
}
The functional margin is not normalized, meaning that its value depends on the scale of the weight vector $\boldsymbol{w}$.
For example, if we multiply $\boldsymbol{w}$ and $w_0$ by a constant factor $\alpha > 1$, 
the functional margin will also increase by the same factor, even though the decision boundary remains unchanged.
Without normalization, we could arbitrarily increase the functional margin by scaling the weights, 
which does not reflect a true increase in confidence.
\\\\To address this, we introduce the concept of \textbf{geometric margin}, 
which normalizes the functional margin by the norm of the weight vector.
\definition{Geometric margin}{
    The distance from a point $\boldsymbol{x}$ to the decision boundary $H$ is called \textbf{geometric margin}.
    \begin{equation}\label{eq:geometric_margin}
        r^x = \frac{f(\boldsymbol{x})}{||\boldsymbol{w}||}
    \end{equation}
}

The distance from the origin to the decision boundary is given by:
\begin{equation}
    r_0 = \frac{|w_0|}{||\boldsymbol{w}||}
\end{equation}

\begin{proof}
A generic point $\boldsymbol{x}$ can be expressed by its \textbf{projection} on $H$ plus its \textbf{distance} from $H$ times the vector in that direction:
\[
    \boldsymbol{x} = \boldsymbol{x}^P + r^x \frac{\boldsymbol{w}}{||\boldsymbol{w}||}
\]
where:
\begin{itemize}
    \item $\boldsymbol{x}^P$ is the projection of $\boldsymbol{x}$ on $H$;
    \item $r^x$ is the distance from $\boldsymbol{x}$ to $H$;
    \item $\frac{\boldsymbol{w}}{||\boldsymbol{w}||}$ is the unit vector in the direction of $\boldsymbol{w}$.
\end{itemize}
Substituting this into the discriminant function:
\[
f(\boldsymbol{x}) = \boldsymbol{w}^T \left( \boldsymbol{x}^P + r^x \frac{\boldsymbol{w}}{||\boldsymbol{w}||} \right) + w_0
\]
\[
= \boldsymbol{w}^T \boldsymbol{x}^P + w_0 + r^x w_t \frac{\boldsymbol{w}}{||\boldsymbol{w}||}
\]
We can notice that $\boldsymbol{w}^T \boldsymbol{x}^P + w_0 = 0$ since $\boldsymbol{x}^P$ lies on the decision boundary $H$. Therefore:
\[
f(\boldsymbol{x}) = r^x ||\boldsymbol{w}||
\]
Rearranging this gives us the formula for the geometric margin as in equation \eqref{eq:geometric_margin}.
\end{proof}

\subsection{Multiclass Classification}
In a \textbf{multiclass classification} problem with $K$ classes, we can extend the linear binary classifier by using \textbf{one-vs-all} (OvA) or \textbf{one-vs-one} (OvO) strategies.
\subsubsection{One-vs-All (OvA)}
The idea is to train $K$ binary classifiers, one for each class. Each classifier $f_k(\boldsymbol{x})$ is trained to distinguish class $k$ from all other classes.
This way, we have a model with $k$ \textbf{hyperplanes}, each separating one class from the rest.
At prediction time, we evaluate all $K$ classifiers and assign the input $\boldsymbol{x}$ to the class with the maximum \textbf{functional margin}.
\subsubsection{One-vs-One (OvO)}
The idea is to train a binary classifier for every pair of classes. For $K$ classes, we need to train $\frac{K(K-1)}{2}$ classifiers.
The model predicts a new example $\boldsymbol{x}$ in the class winning the most pairwise comparisons, as in a tournament.
\subsubsection{Onve-vs-One (OvO) vs One-vs-All (OvA)}
\begin{itemize}
    \item OvA requires to train $K$ classifiers, each on the entire dataset;
    \item OvO requires to train $\frac{K(K-1)}{2}$ classifiers, each on a smaller subset of the data (only the samples belonging to the two classes being compared);
\end{itemize}
If the number of examples is high, OvO can be more efficient, since each classifier is trained on a smaller dataset.