\section{Perceptron}
Perceptron are one of the simplest types of artificial neural networks, based on the concept of linear classifiers (\ref{sec:linear_discriminant_functions}).
They are based on the idea of simulating the behavior of a biological neuron.
\subsection{Biological Motivation}
The human brain is composed of densely interconnected neurons. They process information based on the following principles:
\begin{itemize}
    \item Electrochemical reactions allow signals to propagate through the neuron;
    \item Synapses connect neurons together, they can either \textbf{excite} or \textbf{inhibit} the receiving neuron;
    \item When the incoming signals exceed a certain threshold, the neuron \textbf{fires} and transmits the signal to the next neuron.
\end{itemize}
Artificial neural networks aim to mimic this behavior using mathematical models. 
The threshold is modeled using an \textbf{activation function}, while the synapses are represented by \textbf{weights}.

\subsection{Single Neuron Architecture}
A single neuron perceptron is a specific learning model that implements a linear binary classifier. It can be represented as:
\begin{equation}   
    \label{eq:perceptron}
    f(\boldsymbol{x}) = sign(\boldsymbol{w}^T \boldsymbol{x} + w_0)
\end{equation}
The function $f(\boldsymbol{x})$ is simply a \textbf{linear combination} of the input features $\boldsymbol{x}$, with weights $\boldsymbol{w}$
and an activation function (sign function).
\begin{figure}[H]
    \begin{tikzpicture}
        \node[input-node] (bias) at (0, 2) {1};
        \node[left=0.2cm of bias] {\textbf{Bias}};
        
        \node[input-node] (x1) at (0, 1) {$x^1$};
        \node[input-node] (x2) at (0, 0) {$x^2$};
        \node (dots) at (0, -0.8) {$\vdots$};
        \node[input-node] (xm) at (0, -1.8) {$x^m$};
        
        \node[left=0.5cm of x2] (inputs-label) {Inputs};

        % 2. Draw Summation and Activation Nodes
        \node[sum-node] (sum) at (4, 0.4) {\Large $\Sigma$};
        \node[activation-node, right=1.2cm of sum] (act) {
            \begin{tikzpicture}[scale=0.2] 
                \draw[blue, thick] (-1,-1) -- (0,-1) -- (0,1) -- (1,1); 
            \end{tikzpicture}
        };
        % Connection
        \draw[connection] (bias) -- (sum) node[midway, above] {$w_0$};
        \draw[connection] (x1) -- (sum) node[midway, above] {$w_1$};
        \draw[connection] (x2) -- (sum) node[midway, above] {$w_2$};
        \draw[connection] (xm) -- (sum) node[midway, above] {$w_m$};
        \draw[connection] (sum) -- (act);
        \draw[connection] (act) -- ++(2,0) node[right] {Output};

        \node[above=0.3cm of act] {Activation func};

    \end{tikzpicture}
\end{figure}
\subsubsection{Augmented feature / weight vector}
To simplify the notation, we can introduce an \textbf{augmented feature vector}. We can rewrite the 
equation \eqref{eq:perceptron} as:
\begin{equation}
    f(\boldsymbol{x}) = sign(\hat{\boldsymbol{w}}^T \hat{\boldsymbol{x}})
\end{equation}
where the \textbf{bias term} is included in the weight vector and the feature vector as follows:
\[
    \hat{\boldsymbol{w}} = 
    \begin{bmatrix}
    w_0 \\
    \boldsymbol{W}
    \end{bmatrix}
    , \quad
    \hat{\boldsymbol{x}} = 
    \begin{bmatrix}
    1 \\
    \boldsymbol{x}
    \end{bmatrix}
\]
In the following sections, we will skip the hat notation for simplicity.

\subsubsection{Representational power}
The single neuron perceptron can only represent \textbf{linearly separable} functions. 
Examples of this are the \textbf{primitive boolean functions} such as \textbf{AND}, \textbf{OR}, and \textbf{NOT}.
This implies that any logic formula can be represented by a network of a 2 level perceptron in \textbf{disjunctive normal form} (DNF) or \textbf{conjunctive normal form} (CNF).
For example, the DNF formula:
\[(A \land B) \lor (C \land D)\]
The first layer of the network computes the conjunctions $A \land B$ and $C \land D$ using two perceptrons,
while the second layer computes the disjunction of their outputs using a third perceptron.
\\However, this representation require an \textbf{exponential} number of perceptrons in the worst case.

\subsection{Parameter learning}
Similar to what we saw for \textbf{maximum likelihood estimation}, we need to find a \textbf{function} of the parameters
to be optimized.
\\\\In this case, a reasonable choice is to minimize the \textbf{measure of error} on the training set $\mathcal{D}$. 
This is called \textbf{loss function} $\ell$ and it compares the predicted output $f(\boldsymbol{x}_i)$ 
with the ground truth $y_i$.
\\We can define \textbf{training error} $E$ as:
\begin{equation}
    \label{eq:misclassification_error}
    E(\boldsymbol{w}, D) = \sum_{(\boldsymbol{x},y) \in \mathcal{D}} \ell(f(\boldsymbol{x}), y)
\end{equation}
This is simply the sum of the losses over all training examples.

\subsubsection{Gradient descent}
This is an optimization problem where we want to find the weights $\boldsymbol{w}$ that minimize the error function $E$.
There is no closed-form solution for this problem, so we need to use an iterative optimization algorithm.
\\The common approach is to minimize the error function using a \textbf{gradient descent} approach:
\begin{enumerate}
    \item Initialize the weights $\boldsymbol{w}$ randomly or with zeros.
    \item Iterate until gradient is approximately zero, updating the weights as:
    \begin{equation}
        \boldsymbol{w}' = \boldsymbol{w} - \eta \nabla E(\boldsymbol{w}, \mathcal{D})
    \end{equation}
    where $\eta$ is the learning rate and controls the amount of movement at each step.
\end{enumerate}
Since the gradient $\nabla E(\boldsymbol{w}; D)$ points in the direction of \textbf{maximum increase} of the error function,
we multiply it by $-1$ to move in the direction of \textbf{maximum decrease}.
This approach is guaranteed to converge to a local optimum of the error function $E(\boldsymbol{w}; D)$ for small enough learning rates $\eta$.
\begin{itemize}
    \item too \textbf{low} learning rate: slow convergence;
    \item too \textbf{high} learning rate: the algorithm may \textbf{diverge} (oscillate around the minimum).
\end{itemize}
There are techniques to adaptively change the learning rate during training, allowing for faster convergence.

\subsubsection*{Perceptron training rule}
In order to apply a \textbf{gradient descent} approach, we need to define a \textbf{loss function} which is \textbf{differentiable} (smooth).
For example, the misclassification error defined in \eqref{eq:misclassification_error} 
is not differentiable, as it is piecewise constant.
\\As a consequence, a different training rule is adopted. It's defined as follows:
\begin{equation}\label{eq:perceptron_error}
    E(\boldsymbol{w}; \mathcal{D}) = \sum_{(\boldsymbol{x}, y) \in \mathcal{D}_E} -yf(\boldsymbol{x})
\end{equation}
where $\mathcal{D}_E$ is the set of \textbf{misclassified} samples in the training set $\mathcal{D}$, 
for which it holds that:
\[ yf(\boldsymbol{x}) \leq 0 \]
The error is the \textbf{sum} of the \textbf{functional margins} (confidences) of the misclassified samples.
\\Applying gradient descent to the error function in \eqref{eq:perceptron_error}, we obtain:
\[\nabla E(\boldsymbol{w}; \mathcal{D}) = \nabla \sum_{(\boldsymbol{x}, y) \in \mathcal{D}_E} -yf(\boldsymbol{x})\]
We can substitute $f(\boldsymbol{x})$ with the linear function $\boldsymbol{w}^T \boldsymbol{x}$:
\[=\nabla \sum_{(\boldsymbol{x}, y) \in \mathcal{D}_E} -y(\boldsymbol{w}^T \boldsymbol{x}) \]
The derivative with respect to $\boldsymbol{w}$ is:
\[= \sum_{(\boldsymbol{x}, y) \in \mathcal{D}_E} -y \boldsymbol{x} \]
Thus, the amount of change in the weights is:
\[- \eta \nabla E(\boldsymbol{w}; \mathcal{D}) = \eta \sum_{(\boldsymbol{x}, y) \in \mathcal{D}_E} y \boldsymbol{x} \]
\subsubsection{Stochastic Gradient Descent}
In practice, the \textbf{batch gradient descent} approach we discussed so far is rarely used, 
since it requires to compute the gradient over the entire training set $\mathcal{D}$ at each iteration.
\\A more common approach is to use \textbf{stochastic gradient descent} (SGD). It works as follows:
\begin{enumerate}
    \item Initialize the weights $\boldsymbol{w}$ randomly;
    \item Iterate until all examples are correctly classified:
    \begin{enumerate}
        \item For each incorrectly classified example $(\boldsymbol{x}, y)$ in the training set $\mathcal{D}$, update the weights as:
        \begin{equation}
            \boldsymbol{w} = \boldsymbol{w} + \eta y \boldsymbol{x}
        \end{equation}
    \end{enumerate}
\end{enumerate} 
With this approach, the weights are updated after each misclassified example, 
rather than after processing the entire training set.
Each gradient step is \textbf{very fast} and can sometimes avoid local minima.

\subsection{Perceptron Regression}
Linear models can also be used for \textbf{regression} tasks, where the goal is to predict a continuous output variable $y$ given an input feature vector $\boldsymbol{x}$.
The problem can be modeled as:
\begin{itemize}
    \item Let $X \in \mathbb{R}^{n \times d}$ be the input training matrix (i.e. $X = \left[ \boldsymbol{x}_1, \boldsymbol{x}_2, \ldots, \boldsymbol{x}_n \right]^T$), where each row is a $d$-dimensional feature vector and $n = |\mathcal{D}|$;
    \item Let $\boldsymbol{y} \in \mathbb{R}^n$ be the output training Matrix (i.e. $\boldsymbol{y}_i$ is the target value for the i-th training example);
    \item \textbf{Regression learning} can be stated as a set of \textbf{linear equations}:
    \begin{equation*}
        f(x) = y \implies X\boldsymbol{w} = \boldsymbol{y}
    \end{equation*}
    Giving as a solution:
    \begin{equation}
        \boldsymbol{w} = X^{-1}\boldsymbol{y}
    \end{equation}
\end{itemize}
However this approach has a few problems:
\begin{itemize}
    \item The matrix $X$ is usually not square, so the inverse $X^{-1}$ is not defined;
    \item System of equations is overdetermined (more equations than unknowns) $\implies$ no exact solution exists.
\end{itemize}

\subsubsection{Error function}
We can again resort to loss minimization to find an approximate solution.
Our goal is not to solve the system exactly, but to find the weights $\boldsymbol{w}$ that minimize the \textbf{mean squared error} (MSE) on the training set:
\begin{equation*}
    E(\boldsymbol{w}; \mathcal{D}) = \sum_{(\boldsymbol{X}, y) \in \mathcal{D}} (y - f(\boldsymbol{x}))^2
\end{equation*}
This can be rewritten in matrix form as:
\begin{equation}
    \label{eq:mse}
    E(\boldsymbol{w}; \mathcal{D}) = (\boldsymbol{y} - X\boldsymbol{w})^T (\boldsymbol{y} - X\boldsymbol{w})
\end{equation}
This is a \textbf{convex} function, so it has a single global minimum with a closed-form solution. 
\textbf{Gradient descent} can still be faster than computing the closed-form solution, especially for large datasets.

\subsubsection{Closed form solution}
To find the weights $\boldsymbol{w}$ that minimize the mean squared error, we can set the gradient 
of the error function (\ref{eq:mse}) to zero and solve for $\boldsymbol{w}$:
\[\nabla E(\boldsymbol{w}; \mathcal{D}) = \nabla (\boldsymbol{y} - X\boldsymbol{w})^T (\boldsymbol{y} - X\boldsymbol{w})\]
\[ 2(\boldsymbol{y} - X\boldsymbol{w})^T (-X) = 0\]
\[-2\boldsymbol{y}^TX + 2\boldsymbol{w}^T X^T X = 0\]
Rearranging the terms and dividing by 2, we get:
\[\boldsymbol{w}^T X^T X = \boldsymbol{y}^T X\]
Taking the transpose of both sides:
\[X^T X \boldsymbol{w} = X^T \boldsymbol{y}\]
Assuming that $X^T X$ is invertible, we can solve for $\boldsymbol{w}$:
\begin{equation}
    \boldsymbol{w} = (X^T X)^{-1} X^T \boldsymbol{y}
\end{equation}
NOTE:
\begin{itemize}
    \item $(X^T X)^{-1} X^T$ is a pseudoinverse called the \textbf{left inverse} of $X$;
    \item If $X$ is square, than the left inverse is equal to the regular inverse $X^{-1}$, and this corresponds to the exact solution of the system of equations;
    \item The left inverse exists only if the columns of $X$ are linearly independent. 
    This means that each feature is independent from all the others.
    If this is not the case, we can use remove redundant features using techniques such as
    \textbf{Principal Component Analysis} (PCA), discussed in the Appendix \ref{sec:PCA}
\end{itemize}

\subsubsection{Gradient descent}
In case of large datasets, computing the closed-form solution can be computationally expensive.
Also, we are not always interested in finding the exact minimum of the error function, but rather a good enough approximation. 
\\In such cases, we can use \textbf{gradient descent} to iteratively update the weights until convergence.
Again, given the gradient of the error function, we can compute the update of the weights as:
\[\boldsymbol{w}' = \boldsymbol{w} - \eta \nabla E(\boldsymbol{w}; \mathcal{D})\]
To compute the gradient, we start from the error function defined in \eqref{eq:mse}:
\[ \nabla E(\boldsymbol{w}; \mathcal{D}) = \nabla \sum_{(\boldsymbol{x}, y) \in \mathcal{D}} (y - f(\boldsymbol{x}))^2  \]
We can rewrite the error function in matrix form:
\[ = \nabla (\boldsymbol{y} - X\boldsymbol{w})^2 \] 
\[= \nabla (\boldsymbol{y} - X\boldsymbol{w})^T (\boldsymbol{y} - X\boldsymbol{w})\]
\[ = 2(\boldsymbol{y} - X\boldsymbol{w})^T (-X) \]