\section{Reinforcement Learning}
Reinforcement Learning (RL) is a type of machine learning algorithm where an agent learns to make decisions 
by taking actions in an environment to maximize cumulative rewards.
\subsection{Learning Settings}
\begin{itemize}
    \item The learner (agent) is provided with a set of possible states $S$, and 
    for each state, a set of possible actions $A$, moving it to a new state.
    \item In performing action $a$, from state $s$, the agent receives an immediate reward $r(s,a)$.
    \item The goal is to learn a policy $\pi: S \rightarrow A$ that maximizes the overall reward over time.
    \item The learner has to deal with problems of delayed reward, where the consequences 
    of an action may not be immediately apparent and trade-offs between \textbf{exploration} (trying new actions) 
    and \textbf{exploitation} (choosing actions known to yield high rewards).
\end{itemize}
The agent needs to take a sequence of decisions, maximizing some utility function over time. The result of a decision
may not be always deterministic.
\\\\More formally, the setting of reinforcement learning can be modeled as a Markov Decision Process (MDP).
\subsection{Markov Decision Process (MDP)}
An MDP is defined by:
\begin{itemize}
    \item A set of \textbf{states} $S$, in which the agent can be;
    \item A set of \textbf{terminal states} $S_{G} \subseteq S$, where the process ends. Might also be empty;
    \item A set of \textbf{actions} $A$ available to the agent;
    \item A \textbf{transition model}, providing the probability of going to state $s'$ when taking action $a$ in state $s$: 
    \[
        P(s'|s,a) \;\; s,s' \in S, a \in A
    \]
    \item A \textbf{reward function} $R(s,a,s')$, providing the immediate reward received after transitioning from state $s$ to state $s'$ due to action $a$.
\end{itemize}
The agent's objective is to find a policy $\pi(s)$ that maximizes an utility function, taking into account the uncertainty in state transitions and rewards.

\subsection{Utilities}
Utilities are defined over \textbf{environment histories}. 
An environment history is a sequence of states $[s_0, s_1, s_2, ...]$, visited by the agent over time.
When calculating the utility, we assume an \textbf{infinite horizon}, meaning the agent will continue 
to interact with the environment indefinitely. 
We also assume \textbf{stationary preferences}, meaning the agent's preferences do not change over time.
\\We can define the utility in two ways:
\begin{itemize}
    \item \textbf{Additive rewards}: values immediate and future rewards equally.
    \[ U([s_0, s_1, s_2, \dots]) = R(s_0) + R(s_1) + R(s_2) + \dots \]
    \item \textbf{Discounted utility}: prefers immediate rewards over future rewards.
    \[ U([s_0, s_1, s_2, \dots]) = R(s_0) + \gamma R(s_1) + \gamma^2 R(s_2) + \dots \]
    where $0 \leq \gamma < 1$ is the discount factor.

\end{itemize}
Here we are assuming that rewards only depend on the current state, not the action taken or the next state.

\subsection{Policy}
A policy $\pi$ is a mapping from states to actions: $\pi: S \rightarrow A$. 
It defines the action $a$ that the agent should take when in a given state $s$.
\paragraph{Expected Utility of a policy}
Given a policy $\pi$, we define the expected utility of $\pi$ as the utility of an environment, taken in expectation over all possible state sequences generated by following policy $\pi$.
In other words, we are multiplying the utility of that history, by the probability of that history occurring under policy $\pi$.
\definition{Optimal Policy}{
An optimal policy $\pi^*$ is a policy that maximizes the expected utility. In other words, it is the best policy the agent can follow to achieve the highest cumulative reward over time.
}
\definition{Utility of states}{
    Given a policy $\pi$, the utility of a state $s$ is defined as:
    \[
        U^{\pi}(s) = \mathbb{E}_{\pi} \left[ \sum_{t = 0}^{\inf} \gamma^t R(S_t)|S_0 = s \right]
    \]
    Where $S_t$ is the state reached after $t$ steps, using policy $\pi$, starting from state $S_0 = s$.
    This is the expected utility of the environment history, starting from state $s$ and following policy $\pi$, 
    obtained by summing the discounted rewards over time.
}
The \textbf{true utility} of a state $s$ is defined as the utility of that state under the optimal policy:
\begin{equation}
    \label{trueutility}
    U^*(s) = U^{\pi^*}(s)
\end{equation}
Given the true utility of each state, we can easily derive the optimal policy by choosing the action that maximizes the expected utility:
\label{optpolicy}
\begin{equation}
    \pi^*(s) = \arg\max_{a \in A} \sum_{s' \in S} P(s'|s,a) U^*(s')
\end{equation}
This definition creates a circular dependency between the optimal policy and the true utility, which can be resolved using the Bellman equations.

\subsubsection{Bellman Equations}
The Bellman equation for state $s$ is defined as:
\[ U(s) = R(s) + \gamma \max_{a \in A} \sum_{s' \in S} P(s'|s,a) U(s') \]
This equation states that the utility of a state $s$ is equal to the immediate reward $R(s)$ obtained by being in that state,
plus the discounted expected utility of the next state, assuming the agent takes the optimal action.
\\\\We can define a system of Bellman equations, one for each state in the MDP. 
Solving this system will yield the true utilities for all states, which can then be used to derive the optimal policy.
Directly solving this system can be computationally expensive, since they are non-linear equations (due to the max operator).
Instead, we can use iterative methods such as \textit{value iteration} or \textit{policy iteration} to approximate the solution.

\subsubsection{Value Iteration}
Value iteration, also known as Utility Iteration, works by initializing the utility of all states to arbitrary values (for example zero) and then repeatedly updating the utilities using the Bellman equation until convergence.
\begin{algorithm}[H]
    \caption{Value Iteration}
    \begin{algorithmic}[1]
        \State Initialize $U_0(s)\leftarrow 0 \;\; \forall s \in S$
        \Repeat
            \State Do bellman update for all states $s$:
            \[
            U_{i+1}(s) \leftarrow R(s) + \gamma \max_{a \in A} \sum_{s' \in S} P(s'|s,a) U_i(s')
            \]
            \State $i \leftarrow i + 1$
        \Until{convergence}
        \State \textbf{return} $U(s)$
    \end{algorithmic}
\end{algorithm}
At each iteration, we update the utility of each state based on the current estimates of the utilities of the successor states. 
This process continues until the utilities converge to stable values.
Since we are assuming that the found utilities are the true utilities, the optimal policy can then be derived 
from the converged utilities using the formula for the optimal policy \ref{optpolicy}.

\subsubsection{Policy Iteration}
Contrary to what happens in value iteration (where the policy is derived from the utilities), 
in policy iteration we start with an arbitrary policy and then iteratively improve it.
\begin{algorithm}[H]
    \caption{Policy Iteration}
    \begin{algorithmic}[1]
        \State Initialize $\pi_0$ randomly
        \Repeat
            \State \textbf{Policy Evaluation}: solve the set of linear equations:
            \[
                U_i(s) = R(s) + \gamma \sum_{s' \in S} P(s'|s, \pi_i(s)) U_i(s') \;\; \forall s \in S
            \]
            where $\pi_i(s)$ is the action taken in state $s$ under policy $\pi_i$.
            \State \textbf{Policy Improvement}: update the policy using the new utilities:
            \[
                \pi_{i+1}(s) = \arg\max_{a \in A} \sum_{s' \in S} P(s'|s,a) U_i(s') \;\; \forall s \in S
            \]
        \Until{convergence}
        \State \textbf{return} $\pi(s)$
    \end{algorithmic}
\end{algorithm}
In the policy evaluation step, we compute the utilities of the states under the current policy by 
solving a system of linear equations. 
Using the current policy, instead of the max operator from the Bellman equation,
we made the equations linear, making it easier to solve.
In the policy improvement step, we update the policy using equation \ref{optpolicy}, based on the newly computed utilities.

\subsection{Dealing with Partial Knowledge}
Value iteration and policy iteration both assume that the agent has complete knowledge of the MDP 
(environment, transition model, reward function).
In most real-world scenarios, the agent does not have access to the full transition model or reward function, 
but it aims to learn them through interaction with the environment.
\\There are two main approaches to deal with this partial knowledge:
\begin{itemize}
    \item \textbf{Policy Evaluation}: the policy is given, environment is learned through experience (passive agent). 
    The agent simply learns the transition model while following a fixed policy.
    \item \textbf{Policy Improvement}: both the policy and the environment are learned through experience (active agent). 
\end{itemize}

\subsection{Policy Evaluation}
\label{policyevaluation}
\subsubsection{Adaptive Dynamic Programming (ADP)}
In ADP, the agent maintains counts of useful transitions to estimate the transition model:
\begin{itemize}
    \item $N_{sa}$: number of times action $a$ has been taken in state $s$;
    \item $N_{s'|sa}$: number of times action $a$ has led to state $s'$ from state $s$.
\end{itemize}
\begin{algorithm}[H]
    \caption{Adaptive Dynamic Programming}
    \begin{algorithmic}
        \Repeat
            \State Initialize $s$
            \Repeat
                \State receive reward $r$, set R($s$) = r
                \State choose action $a \leftarrow \pi(s)$
                \State execute action $a$, observe next state $s'$
                \State update counts:
                \[
                    N_{sa} \leftarrow N_{sa} + 1
                \]
                \[
                    N_{s'|sa} \leftarrow N_{s'|sa} + 1
                \]
                \State update transition model:
                \[
                    P(s''|s,a) = \frac{N_{s''|sa}}{N_{sa}} \;\; \forall s'', s \in S, a \in A
                    \]
                \State update utilities using policy evaluation
                \State $U \leftarrow \text{PolicyEvaluation}(\pi,U, p, R, \gamma)$
            \Until{$s$ is terminal}
        \Until{convergence}
    \end{algorithmic}
\end{algorithm}
\definition{Remark:}{
    The outer loop is needed to ensure any significant results. Each iteration of the outer loop is called an \textbf{Episode}.
}
The algorithm performs maximum likelihood estimation of the transition model based on observed counts. 
It then updates the utilities of each state, based on the current learned transition model, using standard policy Evaluation.
This approach is really expensive, since it requires computing the policy evaluation at every step.

\subsubsection{Temporal-Difference Learning (TD Learning)}
TD Learning tries to approximate the utility of states, without needing to run full policy evaluation at every step. This is done by assuming a deterministic transition model (no real transition model is learned).
If the agent is in state $s$ and takes action $a$, arriving in state $s'$, we assume that $s'$ is the only possible successor of $s$:
\begin{itemize}
    \item If $s'$ was always the successor of $s$, then we can update the utility of $s$ as:
    \[ U(s) = R(s) + \gamma U(s') \]
    \item The temporal-difference updates the utility of $s$ incrementally towards this target at each step:
    \[ U(s) \leftarrow U(s) + \alpha \left( R(s) + \gamma U(s') - U(s)\right) \]
\end{itemize}
Where $\alpha$ is the learning rate, controlling how much new information overrides old information.
It should be decreasing over time, to ensure convergence.
\\Basically, the idea is to adjust the utility of the current state $s$, making it closer to the approximation
of the utility based on the observed transition to $s'$.
\begin{algorithm}[H]
    \caption{TD Learning}
    \begin{algorithmic}
        \Repeat
            \State Initialize $s$
            \Repeat
                \State Receive reward $r$
                \State Choose next action $a \leftarrow \pi(s)$
                \State Execute action $a$, observe next state $s'$
                \State update local utility estimates:
                \[
                    U(s) \leftarrow U(s) + \alpha \left( r + \gamma U(s') - U(s)\right)
                \]
                \State $s \leftarrow s'$
            \Until{$s$ is terminal}
        \Until{convergence}
    \end{algorithmic}
\end{algorithm}
\subsubsection{Differences}
\begin{itemize}
    \item TD Learning does not learn a transition model (simply estimates the utility), while ADP does (it uses maximum likelihood estimation);
    \item Each step of TD Learning is computationally cheaper than ADP, since it does not require running policy evaluation;
    \item TD Learning takes longer to converge than ADP.
\end{itemize}
We can say that TD Learning is a rough approximation of ADP.

\subsection{Policy Improvement}
As we have said before, in policy improvement both the optimal policy and the environment are learned through experience.
\subsubsection{A simple solution}
\label{simplepolicyimprovement}
A simple solution to this problem consists in using a variation of ADP, where the policy evaluation is replaced with the computation of the 
optimal policy (both value and policy iteration can be used).
\[ U(s) = R(s) + \gamma \max_{a \in A} \sum_{s' \in S} P(s'|s,a) U(s') \]
This approach theoretically works, but it has two main issues:
\begin{enumerate}
    \item It is computationally expensive, since it requires computing the optimal policy at every step;
    \item This is a greedy approach, which may lead to suboptimal policies. The space is not explored enough, and the agent tend to overfit to the current knowledge of the environment.
\end{enumerate}
To resolve the exploration vs exploitation trade-off, non entirely greedy algorithms should be defined, to encourage exploration of the state space.

\subsubsection{Trade off strategies}
There are two main strategies to deal with the exploration vs exploitation trade-off:
\begin{itemize}
    \item \textbf{$\epsilon$-greedy}: with probability $1 - \epsilon$, the agent chooses the action that maximizes the expected utility (the greedy action). With probability $\epsilon$, the agent chooses a random action (exploration).
    \item \textbf{Define a new Bellman equation}: assign \textbf{higher utilities estimates} to less explored states, to encourage the agent to visit them.
    \[ U^+(s) = R(s) + \gamma \max_{a \in A} f(\sum_{s' \in S} P(s'|s,a) U^+(s'), N_{sa}) \]
    Where the function $f(u,n)$ is directly proportional to $u$ and inversely proportional to $n$ 
    (the number of times action $a$ has been taken in state $s$).
\end{itemize}
Both these strategies help with the suboptimal policies issue, but still require computing the optimal 
policy at every step, which is computationally expensive.

\subsubsection{Utilities of an action:}
To resolve the computational issue we could follow the idea used in section \ref{policyevaluation} and use TD Learning to approximate the utilities of states.
However, TD Learning cannot be directly applied in this context. Here, we are learning both the policy and the environment.
\\\\As we have seen in equation \ref{optpolicy}, the optimal policy depends on the transition model, which is not computed in TD Learning.
\begin{equation}
    \pi^*(s)
    = \arg\max_{a\in A}
    \underbrace{
        \sum_{s' \in S}
        \overbrace{P(s'|s,a)}^{\text{Not computed}}
        \, U^*(s')
    }_{Q(s,a)}
\end{equation}
We could use TD Learning to directly learn an \textbf{action utility function} $Q(s,a)$, which represents the 
utilities of taking action $a$ in state $s$.
\\Given this, we can derive the optimal policy as:
\[ \pi^*(s) = \arg\max_{a \in A} Q(s,a) \]
There are 2 algorithms that use this idea to compute the optimal policy: SARSA (On-policy) and Q-Learning (Off-policy).
\subsubsection*{SARSA (On-policy)}
SARSA (State-Action-Reward-State-Action) is an on-policy TD learning algorithm. It chooses the next action based on a $\epsilon$-greedy policy derived from the current utility Q.
It then takes that action, to reach state $s'$. It uses again the $\epsilon$-greedy policy to choose the next action $a'$ in state $s'$.
\begin{algorithm}[H]
    \caption{SARSA}
    \begin{algorithmic}
        \Repeat
            \State Initialize $s$
            \Repeat
                \State Receive reward $r$
                \State Choose action $a \leftarrow \pi^{\epsilon}(s)$
                \State Execute action $a$, observe reward $r$ and next state $s'$
                \State Choose next action $a' \leftarrow \pi^{\epsilon}(s')$
                \State Update utility estimates:
                \[
                    Q(s,a) \leftarrow Q(s,a) + \alpha \left( r + \gamma Q(s',a') - Q(s,a)\right)
                \]
                \State $s \leftarrow s'$
                \State $a \leftarrow a'$
            \Until{$s$ is terminal}
        \Until{convergence}
    \end{algorithmic}
\end{algorithm}
Since it is an on-policy algorithm, SARSA updates the utility action $Q(s,a)$ using the current policy $\pi^{\epsilon}$. This means that the action $a'$ taken in state $s'$ is also influenced by the exploration strategy, which can lead to more conservative updates.

\subsubsection*{Q-Learning (Off-policy)}
Q-Learning is an off-policy TD learning algorithm. It uses different policies for exploration and for learning the utilities of actions (the greedy policy).
\begin{algorithm}[H]
    \caption{Q-Learning}
    \begin{algorithmic}
        \Repeat
            \State Initialize $s$
            \Repeat
                \State Receive reward $r$
                \State Choose action $a \leftarrow \pi^{\epsilon}(s)$
                \State Execute action $a$, observe reward $r$ and next state $s'$
                \State Update utility estimates:
                \[
                    Q(s,a) \leftarrow Q(s,a) + \alpha \left( r + \gamma \max_{a'} Q(s',a') - Q(s,a)\right)
                \]
                \State $s \leftarrow s'$
            \Until{$s$ is terminal}
        \Until{convergence}
    \end{algorithmic}
\end{algorithm}

\subsection{Scaling to Large Problems}
All the techniques we have seen so far assume a tabular representation of the utility functions. This is not feasible
for large state or action spaces, since the tables would become too large to store and update efficiently.
\\The solution is to use function approximation techniques: we can approximate the utility functions, using parametric functions (such as neural networks).
The problem now shifts to learning the parameters of these functions, instead of maintaining large tables. This can be done using gradient descent methods.
\subsubsection{TD learning: State Utilities}
First, to do gradient descent, we need to define an error function we need to minimize.
\begin{equation}
    E(s, s') = \frac{1}{2} \left( R(s) + \gamma U_\theta(s') - U_\theta(s) \right)^2
\end{equation}
Where $U_\theta(s)$ is the approximated utility function, parameterized by $\theta$. 
This error function measures the distance between the current value function $U_\theta(s)$ and the target value $R(s) + \gamma U_\theta(s')$.
\\We can then compute the gradient of this error function with respect to the parameters $\theta$:
\begin{equation}
    \nabla_\theta E(s, s') = - \left( R(s) + \gamma U_\theta(s') - U_\theta(s) \right) \nabla_\theta U_\theta(s)
\end{equation}
The stochastic gradient descent update rule for the parameters $\theta$ is then given by:
\begin{equation}
    \theta \leftarrow \theta - \alpha \nabla_\theta E(s, s')
\end{equation}
Where $\alpha$ is the learning rate.
\subsubsection{TD learning: Action Utilities}
Similarly, we can define an error function for the action utilities:
\begin{equation}
    E(s, a, s', a') = \frac{1}{2} \left( R(s) + \gamma Q_\theta(s', a') - Q_\theta(s, a) \right)^2
\end{equation}
This error function measures the distance between the current action-value function $Q_\theta(s, a)$ and the target value $R(s) + \gamma Q_\theta(s', a')$.
\\The gradient of this error function with respect to the parameters $\theta$ is:
\begin{equation}
    \nabla_\theta E(s, a, s', a') = - \left( R(s) + \gamma Q_\theta(s', a') - Q_\theta(s, a) \right) \nabla_\theta Q_\theta(s, a)
\end{equation}
The stochastic gradient descent update rule for the parameters $\theta$ is then given by:
\begin{equation}
    \theta \leftarrow \theta + \alpha \nabla_\theta E(s, a, s', a')
\end{equation}
Where $\alpha$ is the learning rate.
